---
title: "COVID-19 and GHG Emissions, What Can We Learn?"
author: "Justin Rhinehart @ jjrhine@uw.edu, Ryan Friesz @ ref7@uw.edu, Tom Jubin @ tomjubin@uw.edu, Liam Dale @ liamdale@uw.edu"
date: "2/18/2022"
output: html_document
---

<!-- 
[]INSERT P01
[]WRITE AUDIENCE PARAGRAPH
[]WRITE SUMMARY PARAGRAPH
[x]LINK CHARTS/TABLES 
-->

<!-- SOURCING FILES -->
```{r, echo = FALSE, warning=FALSE}
source("../source/c01_climate_model.R")
average_percentage_change <- average_percentage_change

source("../source/c02_Euro_avg.R")
# some_figs <- map_variable_here

source("../source/c03_co2_emissions_map.R")
# some_figs <- map_variable_here

```

<!-- VARIABLES!!!
average_percentage_change <- AVG. MODEL EMISSIONS VS REAL WORLD PERCENT DIFF. (14.28% decrease)
German_co2_2020 <- GERMANY CO2 2020
Brazil_c02_2020 <- BRAZIL CO2 2020
Venezuala_co2_2020 <- VENEZUALA CO2 2020
-->

# SUMMARY PARAGRAPH
`r German_co2_2020`I noticed that this was the highest amount of co2 out of the three  countries I chose to look at more closely. I expected that Germany would emit the most since it as seen as the most industrialized out of the three variables. An interesting trend I see with Germany's co2 emissions is that they peaked in 1979, and since then they have been reducing at a significant rate. However between 2019 and 2020, the decrease in co2 was even steeper than normal which I would attribute to policies regarding the Covid 19 pandemic.

`r Brazil_co2_2020` Brazil had significantly less co2 emission than Germany, which makes sense to me because it is not as industrialized. However it is interesting to note that Brazil has over twice the population of Germany, so co2 emissions per capita are a lot higher. Brazil's all time co2 emissions peaked in 2014, and since then they have slowed down. Yet again, there is a very steep drop from 2019 to 2020 in co2 emissions, which shows the affect of the pandemic. However this decrease was more significant in Germany, which could show a difference in government policies.

'r Venezuela_co2_2020' Venezuela had by far the lowest co2 emissions out of the three countries I looked at closely. In addition, the change in emission from 2019 to 2020 was the most dramatic in Venezuela, as total emission decreased by around 15%. This data shows that the pandemic likely caused a lot of co2 emissions to stop in Venezuela. 

# TABLE
```{r}
# source("../source/ghg_table.R")
```
Describe why you included the table, and what information it reveals.

# CHART 1
```{r, echo = FALSE, warning=FALSE}
source("../source/c01_climate_model.R")
ggplotly(line_graph)
```
Describe the purpose of including the chart
Describe any notable observations and insights from the chart

# CHART 2
```{r, echo = FALSE, warning=FALSE}
source("../source/c02_euro_avg.R")
print(bar_graph)
```
Describe the purpose of including the chart
Describe any notable observations and insights from the chart

# CHART 3
```{r, echo = FALSE, warning=FALSE}
source("../source/c03_co2_emissions_map.R")
ggplotly(co2_emissions_map)
```
This chart is intended to show yearly average C02 emissions from the European continent, and their decrease over time, but greater decrease during the pandemic
There is a marked difference in levels of emissions during the pandemic years. While we see a slow decline before the pandemic, much larger year-over-year drops are seen during the pandemic.
