---
title: "COVID-19 and GHG Emissions, What Can We Learn?"
author: "Justin Rhinehart @ jjrhine@uw.edu, Ryan Friesz @ ref7@uw.edu, Tom Jubin @ tomjubin@uw.edu, Liam Dale @ liamdale@uw.edu"
date: "2/18/2022"
output: html_document
---

<!--
[]INSERT P01
[]WRITE AUDIENCE PARAGRAPH
[]WRITE SUMMARY PARAGRAPH
[x]LINK CHARTS/TABLES
-->

<!-- SOURCING FILES -->
```{r, source, echo = FALSE, warning=FALSE, include=FALSE}
source("../source/c01_climate_model.R")
source("../source/c02_Euro_avg.R")
source("../source/c03_co2_emissions_map.R")
```

<!-- VARIABLES!!!
average_percentage_change <- AVG. MODEL EMISSIONS VS REAL WORLD PERCENT DIFF. (14.28% decrease)
German_co2_2020 <- GERMANY CO2 2020
Brazil_c02_2020 <- BRAZIL CO2 2020
Venezuala_co2_2020 <- VENEZUALA CO2 2020
-->

# SUMMARY PARAGRAPH
`r German_co2_2020` I noticed that this was the highest amount of co2 out of the three  countries I chose to look at more closely. I expected that Germany would emit the most since it as seen as the most industrialized out of the three variables. An interesting trend I see with Germany's co2 emissions is that they peaked in 1979, and since then they have been reducing at a significant rate. However between 2019 and 2020, the decrease in co2 was even steeper than normal which I would attribute to policies regarding the Covid 19 pandemic.

`r Brazil_co2_2020` Brazil had significantly less co2 emission than Germany, which makes sense to me because it is not as industrialized. However it is interesting to note that Brazil has over twice the population of Germany, so co2 emissions per capita are a lot higher. Brazil's all time co2 emissions peaked in 2014, and since then they have slowed down. Yet again, there is a very steep drop from 2019 to 2020 in co2 emissions, which shows the affect of the pandemic. However this decrease was more significant in Germany, which could show a difference in government policies.

`r Venezuela_co2_2020` Venezuela had by far the lowest co2 emissions out of the three countries I looked at closely. In addition, the change in emission from 2019 to 2020 was the most dramatic in Venezuela, as total emission decreased by around 15%. This data shows that the pandemic likely caused a lot of co2 emissions to stop in Venezuela.

`r average_percentage_change`

Europes' overall industrial C02 emissions have been in decline for the past 15 years, with an average yearly decrease between 2007 and 2019 of `r DecreaseAvg` Kg. With the outbreak of the pandemic in 2019, we can see the average yearly decline increased to `r Decrease20192020` Kg, a massive `r Diff1920vsAvg` Kg difference from the average up until that point. This helps to display how much the pandemic affected industrial C02 ouput in Europe.

# TABLE
```{r}
# source("../source/ghg_table.R")
```
Describe why you included the table, and what information it reveals.

This table shows the sum of total reported emissions per million metric ton (MMT). The state column includes four states: Florida, Illinois, Texas, and Washington. We specifically chose these states because we wanted to see how the emissions for 2019, which was pre-covid, would compare to 2020, which was post-coivd. As you can see, three of the four states had a significant drop in emissions in the year 2020, but Florida actually had a small increase. This is a unique observation because Florida was widely known to have some of the least strict lockdown rules, which could be the reason their emissions rose while other states were plummeting.

# CHART 1
```{r, echo = FALSE, warning=FALSE}
ggplotly(line_graph)
```
This chart shows CO2 Emissions in Mt/yr for both real world data and the IPCC's climate models. We specifically chose three models, an aggressive (WITCH_AMPERE), a middle of the road (IMAGE 2.4), and a light (GEM-E3-ICCS). The main purpose of including this graph is to demonstrate that GHG emissions actually decreased a notable amount of the pandemic, which is demonstrated by models that predicted what GHG emissions should have looked like if the pandemic hadn't happened.

The main insight we can gather from this chart is that GHG emissions did decrease a significant amount from what we expected, roughly 14.28% less CO2 was released into the atmosphere if we compare an average of those three models to real world data.

# CHART 2
```{r, echo = FALSE, warning=FALSE}
bar_graph
```
This chart is intended to show yearly average C02 emissions from the European continent, and their decrease over time, but greater decrease during the pandemic
There is a marked difference in levels of emissions during the pandemic years. While we see a slow decline before the pandemic, much larger year-over-year drops are seen during the pandemic.


# CHART 3
```{r, echo = FALSE, warning=FALSE}
ggplotly(co2_emissions_map)
```
Describe the purpose of including the chart
Describe any notable observations and insights from the chart

# CHART 4
```{r, echo = FALSE, warning=FALSE}

```
Describe the purpose of including the chart
Describe any notable observations and insights from the chart
